Predicting computational reproducibility of data analysis pipelines \\in large population studies using collaborative filtering

### 2018-07-30 paper.tex
Typos
Improved discussion

% new commands
\newcommand{\todo}[1]{\marginpar{\parbox{18mm}{\flushleft\tiny\color{red}\textbf{TODO}:
      #1}}}
\newcommand{\comment}[1]{\marginpar{\parbox{18mm}{\flushleft\tiny\color{blue}\textbf{Comment}:
      #1}}}

\newcommand{\note}[1]{
  \color{blue}\emph{[Note: #1]}
  \color{black}
}

@@

This paper makes the following contributions:
\begin{enumerate}
\item We model reproducibility evaluations in data processing pipelines as a collaborative filtering problem.
\item We propose strategies to sample the training set under time constraints.
\item We evaluate and compare our sampling strategies on synthetic and real datasets.
\end{enumerate}
The problem formulation, collaborative filtering technique used, and 
proposed sampling strategies for the training set are presented in 
Section~\ref{sec:methods}. The datasets are described in 
Section~\ref{sec:datasets}, and experimental results are reported in 
Section~\ref{sec:results}.
The paper ends with a discussion about the impact, limitations and 
generalizability of our results.

@@

However, only 3 methods are able to provide accuracy values above 0.85 
for all subject types: Random Subjects (RS), RFNU and RFNTL. This is 
due to the fact that these methods are the only ones that can sample 
files produced toward the end of the execution while maintaining a 
reduced number of empty columns. Surprisingly, RFNTS does not perform 
well for more than 2 subject types, even for a training ratio of 0.9. \todo{Check that.}

RS, RFNU and RFNTL achieve a 90\% accuracy for a training ratio of 
about 0.85 \todo{precise that}. Therefore, 15\% of the computations could potentially be 
saved in such reproducibility studies.

@@

\subsection{Varying the number of factors}
Figure~\ref{fig:factors} shows the effect of the number of factors used 
in ALS, for the RFNU method and the synthetic dataset. We do not find a 
clear effect of the number of factors on the accuracy.

\begin{figure}
\includegraphics[width=\columnwidth]{data/results/means_of_results/ALS/Synthetic/8type-RFNU-Differ-Num-Factors/ALS-8-types-ranks.pdf}
\caption{Effect of the number of factors used in ALS, RFNU method, synthetic dataset.}
\label{fig:factors}
\end{figure}

\section{Discussion}

% Which method is the best
While collaborative filtering, perhaps unsurprisingly, is able to 
correctly predict the missing values in a matrix modeling the 
reproducibility problem, the usual random sampling method cannot be 
used in time-constrained processes. We proposed 6 sampling methods to 
address this issue, and we found that one of them, RFNU, \todo{check 
that} is consistently able to solve the problem. We explain that by 
the fact that RFNU leads the training set to contain a balanced mix 
between nearly complete and nearly empty columns, with a continuum of 
intermediate configurations in between. On the contrary, other sampling 
methods, including RFNTL and RFNTS, bias the sampling toward complete 
or empty columns, which is sometimes detrimental to accuracy.

Figure~\ref{fig:overlap} illustrates the comparison between RFNU, RFNTL 
and RFNTS. The matrices in this Figure represent the union between the 
training and the test set, where the training set is represented in 
black (negatives) and white (positives), and the test set is 
represented in green (true positives), yellow (false negatives), gray 
(true negatives) and red (false positives). This representation 
provides insights regarding where, and perhaps why, prediction errors 
occur. At this training ratio, RFNU (Figure~\ref{fig:overlap-rfnu}) 
uniformly samples the number of files per subject between 0.8$N_f$ and 
$N_f$. It provides an opportunity for ALS to be trained on the last 
files of the pipelines, while maintaining a low number of columns with 
a low training ratio. On the contrary, RFNTL 
(Figure~\ref{fig:overlap-rfntl}) samples the number of files per 
subject from 0.85$N_f$, but the probability to have a complete column 
is 0 (the one complete column in Figure~\ref{fig:overlap-rfntl} is the 
one included to prevent cold start issues), which leads to a "stripe" 
of errors at the bottom of the matrix. RFNTS 
(Figure~\ref{fig:overlap-rfnts}) doesn't have this issue, as it 
includes many complete columns in the training set, but this comes at 
the cost of several columns with a very low number of files, lower than 
60\%. For such columns, the prediction may go somewhat wrong, which explains
the reduced accuracy compared to RFNU.

@@

% Training ratio
On datasets that are strongly dependent on row bias, such as the real 
datasets studied here, RFNU provides accuracy values that are 
consistently higher than 0.95 for training ratios higher than 0.5, even 
when biases are not included in the collaborative filtering 
optimization. On the synthetic dataset, RFNU still performs very well 
when biases are not included. However, the training ratio required to 
reach a 0.95 accuracy increases with the number of subject types 
present in the dataset. \todo{maybe find a relation between the 
required training ratio and the number of subjects per subject type in 
the entire dataset}

@@

In any case, we recommend not to include biases in the collaborative filtering 
optimization to solve this problem. Even though biases provide a slight 
accuracy improvement when datasets are strongly biased (constant lines 
in the matrix), they can also be very detrimental for more complex 
datasets such as the synthetic one used here.

% How can this be used in practice
From a practical standpoint, our study shows that reproducibility 
evaluations can be conducted using only 50\% of the files produced by 
the evaluated pipelines, with an accuracy above 95\%. Potentially, this 
could reduce the computing time and storage required for such studies 
by a factor of 2. However, such studies could not be conducted by 
processing only half of the subjects entirely, which would correspond 
to the Complete Columns sampling method. Instead, the processing of all 
the subjects should be initiated, and it should be terminated in a random
uniform way, assuming that files are uniformly produced throughout the 
execution.

@@

% Other possible examples
The method described in this paper and applied to reproducibility 
studies could be used to predict the other variables controlled by 
hidden, deterministic, hidden, parameterized, time-dependent processes, 
for instance markers of chronicle disease activity. In the 
reproducibility example studied in this paper, the variable of interest 
is the reproducibility of files produced by the processing of a given 
subjects, the process is the evaluated pipeline, it is hidden since we 
consider the pipeline as a black box (we don't inspect its source 
code), it is parametrized as it behaves differently on different input 
datasets and parameters, it is time-dependent as the value of the 
variable of interest varies throughout the course of the pipeline 
execution.

>>>>

% new commands
\newcommand{\todo}[1]{\marginpar{\parbox{18mm}{\flushleft\tiny\color{red}\textbf{TODO}:
      #1}}}
\newcommand{\note}[1]{
  \color{blue}\emph{[Note: #1]}
  \color{black}
}

@@

This paper makes the following contributions:
\begin{enumerate}
\item We model reproducibility evaluations in data processing pipelines as a collaborative filtering problem.
\item We propose strategies to sample the training set under time constraints.
\item We evaluate and compare our sampling strategies on synthetic and real datasets.
\end{enumerate}
The problem formulation, collaborative filtering technique used, and 
proposed sampling strategies for the training set are presented in 
Section~\ref{sec:methods}. The datasets are described in 
Section~\ref{sec:datasets}, and experimental results are reported in 
Section~\ref{sec:results}.
We finally conclude on the best sampling method to use, and the 
potential impacts of the results, and on their limitations and 
generalizability.

@@

However, only 3 methods are able to provide accuracy values above 0.85 
for all subject types: Random Subjects (RS), RFNU and RFNTL. This is 
due to the fact that these methods are the only ones that can sample 
files produced toward the end of the execution while maintaining a 
reduced number of empty columns. Surprisingly, RFNTS does not perform 
well for more than 2 subject types, even for a training ratio of 0.9.

Up to 32 subject types, RFNU achieves a 90\% accuracy for a training 
ratio of about 0.85. Therefore, 15\% of the computations could 
potentially be saved in such reproducibility studies.

@@

\subsection{Varying the number of factors}
Figure~\ref{fig:factors} shows the effect of the number of factors used 
in ALS, for the RFNU method and the synthetic dataset. We do not find a 
clear effect of the number of factors on the accuracy.
\begin{figure}[h]
\includegraphics[width=\columnwidth]{data/results/means_of_results/ALS/Synthetic/8type-RFNU-Differ-Num-Factors/ALS-8-types-ranks.pdf}
\caption{Effect of the number of factors used in ALS, RFNU method, synthetic dataset.}
\label{fig:factors}
\end{figure}

\subsection{Error localization}
Figure~\ref{fig:overlap} illustrates the comparison between RFNU, RFNTL 
and RFNTS. The matrices in this Figure represent the union between the 
training and the test set, where the training set is represented in 
black (negatives) and white (positives), and the test set is 
represented in green (true positives), yellow (false negatives), gray 
(true negatives) and red (false positives). This representation 
provides insights regarding where, and perhaps why, prediction errors 
occur. At this training ratio, RFNU (Figure~\ref{fig:overlap-rfnu}) 
uniformly samples the number of files per subject between 0.8$N_f$ and 
$N_f$, which enables the training on the last 
files of the pipeline, while maintaining a low number of columns with 
a low training ratio. On the contrary, RFNTL 
(Figure~\ref{fig:overlap-rfntl}) samples the number of files per 
subject from 0.85$N_f$, but the probability to have a complete column 
is 0 (the one complete column in Figure~\ref{fig:overlap-rfntl} is the 
one included to prevent cold start issues), which leads to a ``stripe" 
of errors at the bottom of the matrix. RFNTS 
(Figure~\ref{fig:overlap-rfnts}) doesn't have this issue, as it 
includes many complete columns in the training set. However, it comes at 
the cost of several columns with a number of files lower than 
60\%: in such columns, the prediction is often entirely wrong, which explains
the reduced accuracy compared to RFNU.

@@

@@

\section{Conclusion}

% Which method is the best
While collaborative filtering, perhaps unsurprisingly, is able to 
correctly predict the missing values in a matrix modeling the 
reproducibility problem, the usual random sampling method cannot be 
used in time-constrained processes. We proposed 6 sampling methods to 
address this issue, and we found that one of them, RFNU, performs 
better than the other ones on average. We 
explain that by the fact that RFNU builds the training set using a 
balanced mix of nearly complete and nearly empty columns, with a 
continuum of 
intermediate configurations. On the contrary, other sampling 
methods, including RFNTL and RFNTS, bias the sampling toward complete 
or empty columns, which is sometimes detrimental to accuracy.

For datasets that are strongly dependent on row bias, such as the 
PreFreesurfer and Freesurfer datasets studied here, RFNU provides 
accuracy values that are consistently higher than 0.95 for training 
ratios higher than 0.5, even 
when biases are not included in the collaborative filtering 
optimization. For the synthetic dataset, RFNU still performs very well 
when biases are not included, but it does very poorly when biases are included.
For this reason, we recommend not to include biases in the collaborative filtering 
optimization to solve this problem. Even though biases provide a slight 
accuracy improvement when datasets are strongly biased (constant lines 
in the matrix), they can also be very detrimental for more complex 
datasets such as the synthetic one used here.

From a practical standpoint, our study shows that reproducibility 
evaluations of the PreFreesurfer and Freesurfer pipelines can be 
conducted using only 50\% of the files produced by the evaluated 
pipelines, with an accuracy above 95\%. Potentially, this could reduce 
the computing time and storage required for such studies by a factor of 
2. However, such studies could not be conducted by processing only half 
of the subjects entirely, which would correspond to the Complete 
Columns sampling method. Instead, the processing of all the subjects 
should be initiated, and it should be terminated in a random uniform 
way, assuming that files are uniformly produced throughout the 
execution. On more complex matrices, such as the synthetic dataset 
studied here, the training ratio required to get a 95\% accuracy 
increases to 0.85 to 0.9.

@@

% Other possible examples
The variable of interest in our study was the reproducibility of files 
produced by the processing of a given subject, and it was controlled by 
a process, the evaluated pipeline. This process was hidden, since we 
considered the pipeline as a black box (we did not inspect its source 
code), it was deterministic, since it was implemented as a program, it 
was parametrized, as it behaved differently on different input datasets 
and parameters, and it was time-dependent, as the value of the 
variable of interest varied throughout the course of the pipeline 
execution. The method described in this paper could possibly be used to 
predict other variables controlled by such hidden, deterministic, 
parametrized, and time-dependent processes, for instance markers of 
chronic disease activity.
